Exploration

### 2018-02-23 exploration.Rmd
env data for 5km NE; IPM generative process only

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pkgs <- c("tidyverse", "magrittr", "fields",
          "maxent", "gbPopMod", "IPMpack", "ppmlasso")
suppressMessages(invisible(lapply(pkgs, library, character.only=TRUE)))
theme_set(theme_bw())
```

@@

## Species description: Life history, characteristics, relevant attributes
The simulated species is fully described here. The object `sp_i` is a list containing three elements: one for each SDM approach. Each of these contains elements for the global parameters, relationships with environmental variables, and demographic matrices that are applicable or potentially applicable for that modeling strategy.
```{r sp_i}
set.seed(1)
sp_i <- list(c_i=list(),
             m1_i=list(),
             m2_i=list())
```

## Landscape
The landscape characteristics are aggregated here. It is a gridded landscape with climate and LULC data.
```{r landscape}
n.rows <- 10
n.cols <- 10
n.cell <- n.rows * n.cols
env.df <- tibble(x=rep(1:n.cols, each=n.rows),
                 y=rep(1:n.rows, times=n.cols),
                 x_y=paste(x, y, sep="_"),
                 temp_mean=rnorm(n.cell),
                 temp_min=rnorm(n.cell),
                 temp_max=rnorm(n.cell),
                 prec_mean=rnorm(n.cell),
                 prec_min=rnorm(n.cell),
                 prec_max=rnorm(n.cell)) %>%
  full_join(.[,3:9], by="x_y", suffix=c("", "_sq")) %>%
  mutate_at(10:15, function(x) x*x)
```

## Correlative: MaxEnt
```{r MaxEnt}

@@

## Mechanistic: CA
```{r CA}

@@

## Mechanistic: IPM
```{r IPM}

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pkgs <- c("fields", "sevcheck", "gbPopMod", "IPMpack", "tidyverse", "magrittr")
suppressMessages(invisible(lapply(pkgs, library, character.only=TRUE)))
theme_set(theme_bw())
lc.df <- read_csv("data/landcover_5km.csv") %>% 
  filter(!is.na(bio1_mean))
```

@@

## Species description: Life history, characteristics, relevant attributes
The simulated species is fully described here. The object `sp_i` is a list containing elements for the global parameters, relationships with environmental variables, and demographic matrices that are applicable or potentially applicable for that modeling strategy. Parameters are abbreviated with the associated kernel component (e.g., `s_` for survival, `g_` for growth, etc) and the type of covariate (`_z` for size, `_x` for environmental covariates).
```{r sp_i}
set.seed(10)
mean_abund <- 600
useLC <- TRUE
regPlots <- TRUE
ipmPlots <- FALSE
n.row <- ifelse(useLC, NA, 10)
n.col <- ifelse(useLC, NA, 20)
n.cell <- ifelse(useLC, n_distinct(lc.df$CellID), n.row*n.col)
n_i <- rpois(n.cell, mean_abund)
sp_i <- list(p=list(s_z=c(-2, 1.3), # b1 + b2*z
                    s_x=c(-.5, -.7, .5, -.1), # b1*x1 + b2*x2 + ...
                    g_z=c(1.5, 2, -0.1), # b1 + b2*z + b3*z^2
                    g_x=c(1, -.5, 2, -.5), #b1*x1 + b2*x2 + ...
                    g_sig2=c(0.01, 0.35), # N(E, s1*exp(2*s2*z))
                    fl_z=c(-6.5, 1.9, -.2, .1), # b1 + b2*z + b3*z^2
                    fl_x=c(0, -.01, 0), # b1*x1 + ...
                    seed_z=c(2.9, 0.05), # b1 + b2*z
                    seed_x=c(-1, -.1, .1, -.2),
                    rcrt=c(1.5, 0.4), #N(mean=rcrt1, sd=rcrt2)
                    p_est=0.02))
```

## Landscape
The landscape characteristics are aggregated here. It is a gridded landscape with climate and LULC data.
```{r landscape}
if(useLC) {
  env.df <- tibble(temp=c(scale(lc.df$bio1_mean)),
                   temp2=temp^2,
                   prec=c(scale(lc.df$bio12_mean)),
                   prec2=prec^2,
                   x=lc.df$left,
                   y=lc.df$top)
} else {
  env.df <- tibble(temp=rnorm(n.cell),
                 temp2=temp^2,
                 prec=rnorm(n.cell),
                 prec2=prec^2,
                 x=rep(1:n.col, each=n.row),
                 y=rep(1:n.row, times=n.col)) 
}
```

---

## Simulation: IPM framework
```{r S}
# set up storage objects
p <- sp_i$p
d <- vector("list", length=n.cell)
lambdas <- rep(NA, n.cell)

# set up environmental covariates
n_sz <- length(p$s_z); n_sx <- length(p$s_x)
n_gz <- length(p$g_z); n_gx <- length(p$g_x)
n_flz <- length(p$fl_z); n_flx <- length(p$fl_x)
n_seedz <- length(p$seed_z); n_seedx <- length(p$seed_x)
X.s <- as.matrix(env.df[,1:n_sx])
X.g <- as.matrix(env.df[,1:n_gx])
X.fl <- as.matrix(env.df[,1:n_flx])
X.seed <- as.matrix(env.df[,1:n_seedx])

# loop through cells
for(i in 1:n.cell) {
  d[[i]] <- data.frame(size=runif(n=n_i[i], min=1, max=7))
  lo <- min(d[[i]]$size)*0.1
  hi <- max(d[[i]]$size)*2
  size.mx <- cbind(1, d[[i]]$size, d[[i]]$size^2, d[[i]]$size^3)

  # Survival
  E_surv <- antilogit(size.mx[,1:n_sz] %*% p$s_z + c(X.s[i,] %*% p$s_x))
  d[[i]]$surv <- rbinom(n_i[i], 1, E_surv)

  # Growth
  E_grow <- size.mx[,1:n_gz] %*% p$g_z + c(X.g[i,] %*% p$g_x)
  d[[i]]$sizeNext <- rnorm(n_i[i], mean=E_grow, 
                           sd=p$g_sig2[1]*exp(2*p$g_sig2[2]*d[[i]]$size))
  d[[i]]$sizeNext <- ifelse(d[[i]]$surv, d[[i]]$sizeNext, NA)

  # Flowering 
  E_fl <- antilogit(size.mx[,1:n_flz] %*% p$fl_z + c(X.fl[i,] %*% p$fl_x))
  d[[i]]$fl <- rbinom(n_i[i], 1, E_fl)

  # Seed production
  E_seed <- d[[i]]$fl * exp((size.mx[,1:n_seedz] %*% p$seed_z + 
                               c(X.seed[i,] %*% p$seed_x)))

  d[[i]]$seed <- rpois(n_i[i], E_seed)
  d[[i]]$seed <- ifelse(d[[i]]$fl, d[[i]]$seed, 0)
  # d[[i]]$fl <- ifelse(d[[i]]$surv, d[[i]]$fl, NA)
  # d[[i]]$seed <- ifelse(d[[i]]$surv, d[[i]]$seed, NA)
  seed.range <- range(d[[i]]$seed, na.rm=TRUE)

  # recruits
  rcrt.n <- max(sum(d[[i]]$seed, na.rm=TRUE)*p$p_est, 1)
  rcrt.i <- (n_i[i]+1):(n_i[i]+rcrt.n)
  d[[i]][rcrt.i,] <- NA
  d[[i]]$sizeNext[rcrt.i] <- rnorm(rcrt.n, p$rcrt[1], p$rcrt[2])

  # Kernel
  so <- makeSurvObj(d[[i]], surv ~ size + size2)
  go <- makeGrowthObj(d[[i]], sizeNext ~ size + size2)
  fo <- makeFecObj(d[[i]], 
                   Formula=c(fl~size+size2+size3, seed~size), 
                   Family=c("binomial", "poisson"),
                   Transform=c("none", "none"))
  P.mx <- makeIPMPmatrix(growObj=go, survObj=so, minSize=lo, maxSize=hi,
                         correction="constant")
  F.mx <- makeIPMFmatrix(fecObj=fo, minSize=lo, maxSize=hi, correction="constant")
  IPM <- P.mx + F.mx
  lambdas[i] <- Re(eigen(IPM)$values[1])

  # Plots
  if(regPlots) {
    par(mfrow=c(3,2), mar=c(4,4,2,1)); main.i <- paste("Cell", i)
    z.t <- d[[i]][!is.na(d[[i]]$size), ]
    plot(surv~size, data=d[[i]], ylim=c(0,1), pch=20, col=rgb(0,0,0,0.2),
         main=paste("Surv:", main.i), xlab="Size (t)", ylab="P(surv to t+1)")
      points(z.t$size, E_surv, cex=0.1, col="red")
    plot(sizeNext~size, data=z.t, pch=20, col=rgb(0,0,0,0.2), 
         main=paste("Growth:", main.i), xlab="Size (t)", ylab="Size (t+1)")
      points(z.t$size, E_grow, cex=0.1, col="red")
      abline(a=0, b=1, col="gray", lty=2)
    plot(fl~size, data=z.t, pch=20, ylim=c(0,1), col=rgb(0,0,0,0.2),
         main=paste("P(Flower:", main.i), xlab="Size (t)", ylab="P(flower)")
      points(z.t$size, E_fl, cex=0.1, col="red")
    plot(seed~size, data=z.t, pch=23-z.t$fl*3, col=rgb(0,0,0,0.2), 
         main=paste("nSeeds:", main.i), xlab="Size (t)", ylab="Number of seeds")
      points(z.t$size, E_seed, cex=0.1, col="red")
    plot(P.mx@meshpoints, Re(eigen(IPM)$vector[,1]), ylab="", type="l", lwd=2,
         main=paste("Stable size distribution:", main.i), xlab="Size (t)")
    hist(d[[i]]$sizeNext[rcrt.i], main=paste("Recruit sizes:", main.i), 
         xlab="Recruit size (t+1)", col="grey", freq=FALSE)
      xx <- seq(0, hi*.5, length.out=100)
      lines(xx, dnorm(xx, p$rcrt[1], p$rcrt[2]), col="red")
  }
  if(ipmPlots) {
    par(mfrow=c(2,2), mar=c(4,4,2,1)); main.i <- paste("Cell", i)
    sens.mx <- sens(IPM); elas.mx <- elas(IPM)
    image.plot(P.mx@meshpoints, P.mx@meshpoints, t(P.mx), 
               main=paste("P:", main.i), xlab="Size (t)", ylab="Size (t+1)")
    image.plot(F.mx@meshpoints, F.mx@meshpoints, t(F.mx),
               main=paste("F:", main.i), xlab="Size (t)", ylab="Size (t+1)")
    image.plot(1:dim(sens.mx)[1], 1:dim(sens.mx)[1], t(sens.mx),
               main=paste("Sens:", main.i), xlab="Size (t)", ylab="Size (t+1)")
    image.plot(1:dim(elas.mx)[1], 1:dim(elas.mx)[1], t(elas.mx),
               main=paste("Elas:", main.i), xlab="Size (t)", ylab="Size (t+1)")
  }
}
lam.df <- env.df %>% add_column(lambda=lambdas)
ggplot(lam.df, aes(x=x, y=y, fill=log(lambdas))) + geom_tile() +
  scale_fill_gradient2(low="darkblue", mid="white", high="darkred", midpoint=0)
par(mfrow=c(2,2))
plot(lambdas ~ temp, data=lam.df, log="y")
plot(lambdas ~ temp2, data=lam.df, log="y")
plot(lambdas ~ prec, data=lam.df, log="y")
plot(lambdas ~ prec2, data=lam.df, log="y")
```

----

## Observed datasets
The simulated species *S* is sampled appropriately in order to fit each SDM model. The observed datasets,*O*, may be random or intentionally biased to mimic common issues with empirical data.
### Correlative: MaxEnt
```{r O_Mx}

```
### Mechanistic: CA
```{r O_CA}

```
### Mechanistic: IPM
```{r O_IPM}

@@

---

## Fitting the models
Each observed dataset *O* is used to predict the distribution of species *S*. The predicted distribution *P* is compared to the known truth *S* to assess the predictive accuracy of each modeling approach, given any limitations, uncertainty, or bias imposed in *O*. 
### Correlative: MaxEnt
```{r P_Mx}

```
### Mechanistic: CA
```{r P_CA}

```
### Mechanistic: IPM
```{r P_IPM}

@@
